### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/1-introduction.tex
Finished review draft, sending to coauthors.

@@

The theory for sufficient and {\it sharp} necessary bounds for
monotone quintic interpolation has been provided [Ulrich and Watson
  1994; He{\ss} and Schmidt 1994] and that theory was recently
utilized in a proposed algorithm [Lux 2020] for monotone quintic
spline construction, however no published mathematical software exists
for the quintic case based on sharp monotonicity conditions. The
software presented here represents the first published software
package for producing $C^2$ shape preserving splines based on sharp
monotonicity conditions. This work improves upon the algorithm
presented by Lux et al. [2020] by refactoring computations for
improved numerical stability, estimating minimum magnitude second
derivatives at breakpoints with a quadratic facet model, and using a
binary search to reduce the magnitude of the modifications made to
initial derivative estimates when constructing a monotone spline
interpolant.

@@

@@

The remainder of this paper is structured as follows: Section 2 provides
the algorithms for constructing a $C^2$ monotone quintic spline interpolant
to monotone data, Section 3 outlines the method of spline representation
($B$-spline basis) and evaluation, Section 4 analyzes the complexity and
sensitivity of the algorithms in {\tt MQSI}, and Section 5 presents 
timing data and some graphs of constructed interpolants.

>>>>

\beginred

@@

\endred

The theory for sufficient and \red{{\it sharp}} necessary bounds for
monotone quintic interpolation has been provided [Ulrich and Watson
  1994; He{\ss} and Schmidt 1994] and that theory was recently
utilized in a proposed algorithm [Lux 2020] for monotone quintic
spline construction, however no published mathematical software exists
for the quintic case based on sharp monotonicity conditions. The
software presented here represents the first published software
package for producing $C^2$ shape preserving splines based on sharp
monotonicity conditions. This work improves upon the algorithm
presented by Lux et al. [2020] by refactoring computations for
improved numerical stability, estimating minimum magnitude second
derivatives at breakpoints with a quadratic facet model, and using a
binary search to reduce the magnitude of the modifications made to
initial derivative estimates when constructing a monotone spline
interpolant.

\beginred

@@

\endred

@@

The remainder of this paper is structured as follows: Section 2 provides
the algorithms for constructing a $C^2$ monotone quintic spline interpolant
to monotone data, Section 3 outlines the method of spline representation
($B$-spline basis) and evaluation, Section 4 analyzes the complexity and
sensitivity of the algorithms in {\tt MQSI}, and Section 5 presents 
timing data and some graphs of constructed interpolants \red{as well
  as a visual comparison with existing monotone spline packages}.

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/4-analysis.tex
Finished review draft, sending to coauthors.

\heading{4. COMPLEXITY AND SENSITIVITY}
Algorithms 1, 2, and 4 have ${\cal O}(1)$ runtime. Given a fixed
schedule for shrinking derivative values, Algorithm 3 has a ${\cal
  O}(n)$ runtime for $n$ data points. In execution, the majority of
the time, still ${\cal O}(n)$, is spent solving the banded linear
system of equations for the B-spline coefficients. Thus for $n$ data
points, the overall execution time is ${\cal O}(n)$.

The quadratic facet model produces a unique sensitivity to input
perturbation, as small changes in input may cause different quadratic
facets to be associated with a breakpoint, and thus different initial
derivative estimates can be produced. This phenomenon is depicted in
Figure 1. Despite this sensitivity, the quadratic facet model is still
preferred because it exactly captures local linear and quadratic
behavior while empirically producing final approximations with less
wiggle ($L^2$ norm of the second derivative) than other methods. A
weighted harmonic mean estimate of first derivatives may be more
accurate when the underlying function changes at a rate greater than a
quadratic, but that method increases the second derivative sensitivity
to small perturbations in data and empirically results in quintic
splines with greater wiggle.

@@

\topinsert
\centerline{\epsfxsize=4truein \epsffile{vis/1-sensitivity.eps}}
{\everymath={\scriptstyle}
\narrower\noindent\rmVIII Fig.\ 1. A demonstration of the quadratic
  facet model's sensitivity to small data perturbations. This example is
  composed of two quadratic functions $f_1(x) = x^2$ over points $\{1$,
  $2$, $5/2\}$, and $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$,
  $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have the same
  curvature. Given the exact five data points seen above, the quadratic
  facet model produces the slope seen in the solid blue line at $x = 5/2$.
  However, by subtracting the value of $f_3$ $= \epsilon(x-2)^2$ from
  points at $x = 3$, 4, where $\epsilon$ is the machine precision
  ($2^{-52}$ for an IEEE 64-bit real), the quadratic facet model produces
  the slope seen in the dashed red line at $x = 5/2$. This is the nature
  of a facet model and a side effect of associating data with local facets.
\par} \endinsert

>>>>

\heading{4. COMPLEXITY AND SENSITIVITY}

Algorithms \red{1 and 4 have ${\cal O}(n)$ runtime for $n$ data
points. Algorithm 2 has a fixed cost ${\cal O}(1)$.} Given a fixed
schedule for shrinking derivative values, Algorithm 3 has a ${\cal
  O}(n)$ runtime for $n$ data points. In execution, the majority of
the time, still ${\cal O}(n)$, is spent solving the banded linear
system of equations for the B-spline coefficients. Thus for $n$ data
points, the overall execution time is ${\cal O}(n)$.

The quadratic facet model produces a unique sensitivity to input
perturbation, as small changes in input may cause different quadratic
facets to be associated with a breakpoint, and thus different initial
derivative estimates can be produced. This phenomenon is depicted in
Figure 1. Despite this sensitivity, the quadratic facet model is still
preferred because it exactly captures local linear and quadratic
behavior while empirically producing final approximations with less
wiggle (\red{local} $L^2$ norm of the second derivative) than other methods. A
weighted harmonic mean estimate of first derivatives may be more
accurate when the underlying function changes at a rate greater than a
quadratic, but that method increases the second derivative sensitivity
to small perturbations in data and empirically results in quintic
splines with greater wiggle.

@@

\topinsert
\centerline{\epsfxsize=4truein \epsffile{vis/1-sensitivity.eps}}
{\everymath={\scriptstyle}

\narrower\noindent\rmVIII Fig.\ 1. A demonstration of the quadratic
  facet model's sensitivity to small data perturbations. This example is
  composed of two quadratic functions $f_1(x) = x^2$ over points $\{1$,
  $2$, $5/2\}$, and $f_2(x) = (x-2)^2 + 6$ over points $\{5/2$, $3$,
  $4\}$. Notably, $f_1(5/2) = f_2(5/2)$ and $f_1$, $f_2$ have the same
  curvature \red{(equal second derivatives)}. Given the exact five data points
  seen above, the quadratic facet model produces the slope seen in the
  solid blue line at $x = 5/2$. However, by subtracting the value of
  $f_3$ $= \epsilon(x-2)^2$ from points at $x = 3$, 4, where $\epsilon$
  is the machine precision ($2^{-52}$ for an IEEE 64-bit real), the
  quadratic facet model produces the slope seen in the dashed red line
  at $x = 5/2$. This is the nature of a facet model and a side effect
  of associating data with local facets.
\par} \endinsert

### 2022-03-11 [2020-08]_ACMTOMS_(MQSI)/paper/5-applications.tex
Finished review draft, sending to coauthors.

@@

\topinsert
\centerline{\epsfxsize=4truein \epsffile{vis/2-piecewise-polynomial.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/3-large-tangent.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/4-signal-decay.eps}}
{\narrower\noindent\rmVIII Fig.\ 2.
{\ttVIII MQSI} results for three of the functions in the included test
suite. The {\itVIII piecewise polynomial} function (top) shows the
interpolant capturing local linear segments, local flats, and
alternating extreme points. The {\itVIII large tangent} (middle)
problem demonstrates outcomes on rapidly changing segments of data.
The {\itVIII signal decay} (bottom) alternates between extreme values
of steadily decreasing magnitude.
\par}
\vskip 6mm
\centerline{\epsfxsize=4truein \epsffile{vis/5-real-data.eps}}
{\narrower\noindent\rmVIII Fig.\ 3.
{\ttVIII MQSI} results when approximating the cumulative distribution
function of system throughput (bytes per second) data for a computer
with a 3.2 GHz CPU performing file read operations from Cameron et al.
[2019]. The empirical distribution of 30 thousand throughput values is
shown in the red dashed line, while the solid line with stylized
markers denotes the approximation made with MQSI given equally spaced
empirical distribution points from a sample of size 100.
\par}
\endinsert

\topinsert
\centerline{\epsfxsize=4truein \epsffile{vis/6-random-monotone.eps}}
{\narrower\noindent\rmVIII  Fig.\ 4.
The {\itVIII random monotone} test poses a particularly challenging
problem with large variations in slope. Notice that despite drastic
shifts in slope, the resulting monotone quintic spline interpolant
provides smooth and reasonable estimates to function values between data.
\par}
\iffalse
\vskip 15mm
\centerline{\epsfxsize=4truein \epsffile{vis/7-runtimes-10.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/8-runtimes-100.eps}}
{\everymath={\scriptstyle} \narrower\noindent\rmVIII Fig.\ 5.
Median total runtime of the routine {\ttVIII MQSI} for number of data points
$n=10$
to 100 (top) and $n=100$ to 1000 (bottom), generated from 100
repeated trials averaged over $14$ different testing functions. The
timings are colored by major algorithmic component, but the vast
majority of execution time is spent solving the banded linear system
of equations in the {\ttVIII FIT\_SPLINE} routine. The runtimes for
the quadratic facet (Algorithm 1) take roughly one microsecond
($10^{-6}$ seconds) per breakpoint, while the binary search
(Algorithm 4) takes roughly four microseconds per breakpoint.
\par}
\fi
\endinsert

@@

Comparisons between this work and lower order methods are problematic,
especially with sparse data. For exact data from smooth functions,
comparing derivative values, higher order methods always win (all
other things being equal) as the data density increases. For sparse or
synthetic data, what constitutes a better fit is either subjective or
dependent on the specific problem. If an application demands $C^2$
continuity and monotonicity, the target of {\tt MQSI}, then this
package provides such interpolants while no existing published spline
software suffices.

>>>>

\topinsert
\textBlack
\centerline{\epsfxsize=4truein \epsffile{vis/2-piecewise-polynomial.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/3-large-tangent.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/4-signal-decay.eps}}
{\narrower\noindent\rmVIII Fig.\ 2.
{\ttVIII MQSI} results for three of the functions in the included test
suite. The {\itVIII piecewise polynomial} function (top) shows the
interpolant capturing local linear segments, local flats, and
alternating extreme points. The {\itVIII large tangent} (middle)
problem demonstrates outcomes on rapidly changing segments of data.
The {\itVIII signal decay} (bottom) alternates between extreme values
of steadily decreasing magnitude.
\par}
\vskip 6mm
\centerline{\epsfxsize=4truein \epsffile{vis/5-real-data.eps}}
{\narrower\noindent\rmVIII Fig.\ 3.
{\ttVIII MQSI} results when approximating the cumulative distribution
function of system throughput (bytes per second) data for a computer
with a 3.2 GHz CPU performing file read operations from Cameron et al.
[2019]. The empirical distribution of 30 thousand throughput values is
shown in the red dashed line, while the solid line with stylized
markers denotes the approximation made with {\tt MQSI} given
equally spaced empirical distribution points from a sample of size 100.
\par}
\endinsert

\topinsert
\centerline{\epsfxsize=4truein \epsffile{vis/6-random-monotone.eps}}
{\narrower\noindent\rmVIII  Fig.\ 4.
The {\itVIII random monotone} test poses a particularly challenging
problem with large variations in slope. Notice that despite drastic
shifts in slope, the resulting monotone quintic spline interpolant
provides smooth and reasonable estimates to function values between data.
\par}
\iffalse
\vskip 15mm
\centerline{\epsfxsize=4truein \epsffile{vis/7-runtimes-10.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/8-runtimes-100.eps}}
{\everymath={\scriptstyle} \narrower\noindent\rmVIII Fig.\ 5.
Median total runtime of the routine {\tt MQSI} for number of data points
$n=10$
to 100 (top) and $n=100$ to 1000 (bottom), generated from 100
repeated trials averaged over $14$ different testing functions. The
timings are colored by major algorithmic component, but the vast
majority of execution time is spent solving the banded linear system
of equations in the {\ttVIII FIT\_SPLINE} routine. The runtimes for
the quadratic facet (Algorithm 1) take roughly one microsecond
($10^{-6}$ seconds) per breakpoint, while the binary search
(Algorithm 4) takes roughly four microseconds per breakpoint.
\par}
\fi
\endinsert

@@

@@

\beginred

The binary search procedure provably converges on the boundary of the
region of monotonicity precisely defined by Ulrich and Watson [1994]
through the application of a boolean function that is guaranteed to be
true at one end of an interval. Since a local quadratic interpolant is
applied for initial function derivative estimates, the approximation
order of the resulting fit is $O(h^3)$ (as for any second order
approximation), but it should be noted that asymptotic approximation
order is rarely of import when considering the scattered sparse data
that $C^2$ approximations like {\tt MQSI} provide. Were exact
evaluations provided, higher order methods always win \textRed (all
other things being equal) as the data density increases. For sparse
data, what constitutes a better fit is either subjective or dependent
on the problem.

Lastly, comparisons between {\tt MQSI} and four published spline
software packages are provided that allow subjective inspection on the
same four test problems as presented above. In each of Figures 5
through 8 the intricacies of the approximation created by {\tt MQSI}
can be contrasted with two monotone $C^1$ quadratic splines (TOMS 574,
Schumaker), the popular $C^1$ cubic method {\tt PCHIP}, and the $C^2$
quintic method {\tt BVSPIS} of TOMS 770. Figures 5 and 8 demonstrate
glaring numerical errors in the Schumaker and {\tt BVSPIS}
implementations, and overall it can be observed that the minimal and
tight conditions on monotonicity provided by TOMS 574 tend to be more
visually appealing than the weaker sufficient conditions utilized by
{\tt PCHIP}. In general the major determining factor for the quality
of {\tt MQSI} is the local quadratic facet model. For applications
where the underlying function is presumed to be a piecewise polynomial
(of relatively low order), the choice of local quadratic interpolants
is reasonable. However the consequence of this decision is that
functions with superquadratic rates of change will tend to have
consistently underestimated first and second derivatives. This
limitation is accepted in favor of perfectly capturing lower order
phenomenon. Were future research to apply alternative initializations
for $C^2$ quintic splines, those could comfortably be made monotone by
the procedures of {\tt MQSI} and most importantly by the application
of these sharp monotonicity conditions.

\endred

If an application demands $C^2$ continuity and monotonicity, the
target of {\tt MQSI}, then this package \red{uniquely} provides such
interpolants \red{based on sharp monotonicity constraints.}

\goodbreak

\topinsert
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/1-toms574.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/2-schumaker.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/3-pchip.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/piecewise-polynomial/4-bvspis.eps}}
{\narrower\noindent\rmVIII Fig.\ 5.
{\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
(second, red), {\tt PCHIP} (third, green), and {\tt BVSPIS} (fourth, purple)
respectively on the {\itVIII piecewise polynomial} test function. {\tt MQSI}
is styled as a gray thin line in the background for comparison. The
Schumaker code (second) fails to produce a monotone approximation for
this problem while producing no errors or warnings. The primary
comparative observation of {\tt MQSI} otherwise is its exact reproduction of
the linear segment between $0.3$ and $0.5$.
\par}
\endinsert

\topinsert
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/1-toms574.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/2-schumaker.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/3-pchip.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/large-tangent/4-bvspis.eps}}
{\narrower\noindent\rmVIII Fig.\ 6.
{\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
(second, red), {\tt PCHIP} (third, green), and {\tt BVSPIS} (fourth, purple)
respectively on the {\itVIII large tangent} test function. {\tt MQSI} is
styled as a gray thin line in the background for comparison. This
problems highlights the core weakness of the quadratic facet model
approach to derivative estimation in {\tt MQSI}, which consistently
underestimates the actual curvature of this function. This weakness is
accepted for its greater accuracy when estimating local lower order
components of functions.
\par}
\endinsert

\topinsert
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/1-toms574.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/2-schumaker.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/3-pchip.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/signal-decay/4-bvspis.eps}}
{\narrower\noindent\rmVIII Fig.\ 7.
{\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
(second, red), {\tt PCHIP} (third, green), and {\tt BVSPIS} (fourth, purple)
respectively on the {\itVIII signal decay} test function. {\tt MQSI} is
styled as a gray thin line in the background for comparison. The most
notable differences between {\tt MQSI} and other approaches can be observed
near local extrema, where {\tt MQSI} produces smaller magnitude second
derivatives.
\par}
\endinsert

\topinsert
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/1-toms574.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/2-schumaker.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/3-pchip.eps}}
\centerline{\epsfxsize=4truein \epsffile{vis/comparisons/random-monotone/4-bvspis.eps}}
{\narrower\noindent\rmVIII Fig.\ 8.
{\ttVIII MQSI} compared with each of TOMS 574 (first, blue), Schumaker
(second, red), {\tt PCHIP} (third, green), and {\tt BVSPIS} (fourth, purple)
respectively on the {\itVIII random monotone} test function. {\tt MQSI} is
styled as a gray thin line in the background for comparison. Notice
that the numerical conditions for this approximation problem are
challenging enough that both the Schumaker and {\tt BVSPIS} codes
incorrectly produce nonmonotone segments (neither of which produced
error codes or any form of indication that a failure
occurred). Notably only {\tt MQSI} and TOMS 574 satisfy tight monotonicity
constraints while also correctly preserving monotonicity in their
approximations. The main segment of divergence between the methods
happens on the $(0.45, 0.5)$ interval, where {\tt MQSI} favors maximizing
the slope on the left side of the interval because the quadratic
interpolant of the right sided points produces an (invalid) negative
slope estimate on the left side of the interval.
\par}
\endinsert
